Response to Pachter’s Review

### 2020-01-22 Response.Rmd

---
title: "Response to Pachter’s Review"
author: "Joshua Paik and Igor Rivin"
date: "1/22/2020"
output:
  pdf_document:
    keep_tex: yes
    toc: yes
  html_document:
    toc: yes
---

>*"Academic politics are so vicious precisely because the stakes are so small."*
> - Henry Kissinger

>*"The value of preprints is in their ability to accelerate research via the rapid dissemination of methods and discoveries."* 
> - [Lior Pachter](https://liorpachter.wordpress.com/2019/10/21/zero-data-rna-seq/)
```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=4) 
```

```{r, echo = FALSE, warning = FALSE, include = FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(MASS)
library(naniar)

#cleaning
df1 <- read_csv("./withAMS.csv")
df1$age <- (2020- df1$year)
df1$citperyear <- df1$citations/df1$age
df1$amscitperyear <- df1$amscit/df1$age
df1$lettergroup <- as.factor(df1$lettergroup)
df1$gender <- as.factor(df1$gender)
df1$role <- as.factor(df1$role)
df1$highered <- as.factor(df1$highered)
df1$institution <- as.factor(df1$institution)
df1$research <- as.factor(df1$research)
df1$country <- as.factor(df1$country)
df1$security <- as.factor(df1$security)
df1$field <- as.factor(df1$field)
df1$simplefield <- as.factor(df1$simplefield)
```
## Introduction

Lior Pachter published a [review](https://liorpachter.wordpress.com/2020/01/17/diversity-matters/) of our [paper](https://arxiv.org/pdf/2001.00670.pdf) claiming that age was the greatest contributor to the observed difference in citations between signers of Letters A, B, and C. Any characterization of our paper which says we do not account for age is *false*. In this response, we will clarify a few points in our paper, repeat the relevant analyses, and show that citations per year is an age agnostic metric to compare mathematicians. We will also go to some length to address potential objections to this new analysis, and also show they are categorically false. **To clarify, when comparing mean and median citations per year amongst R1 Math Professors, A < B < C.** This result still stands, as does the rest of our analysis, post some revision to our data. Finally, we will clearly demonstrate Professor Pachter's mistake, namely parameter tuning (hypertuning) of his arbitrary age cutoff, to acheive a result which supports his incorrect interpretation.

We appreciate Professor Pachter's review - it gives us a chance to make our analysis stronger. We would note that a revision was already in the works and that in a normal review process we would have had three months to respond. However, as our character and ability as scientists were attacked, we thought it was appropriate to reply as quickly as possible.

## Corrections and Clarifications

We would like to thank Lior for finding the bug in our appendix which pushes the mean Google Scholar citations of B further away from A. We agree that the sentence - “while this is not optimal, a quick sample size calculation shows that one needs 303 samples or 21% of the data to produce statistics at a 95% confidence level and a 5% confidence interval.” - is ridiculous.

We should explain exactly how the data collection took place. We used the [scholarly api](https://pypi.org/project/scholarly/) to initially collect our Google Scholar citations data. But the issues were that the scraper did not accurately differentiate between those who had a generic name and the observed fact that older mathematicians (like Cheeger or Gromov) do not have Google Scholar citations. To assure data quality, we manually checked the google scholar citations of every single letter signer, comparing publications when necessary.

However, the empirical difference in citations was staggering and we could predict an objection. More professors from R2 (teaching focused universities) signed A, so it could have pushed the average down. We had already spent so much effort collecting Google Scholar citations, so we made a choice to only collect MathSciNet data on R1 Math Professors, which is why Professor Pachter did not have MathSciNet citations in our dataset. This choice was not made explicitly enough in our first version.  Let us look at the NaN values of those who are full math professors at R1 univesities.

```{r, echo = FALSE}
restrictedData1 <- filter(df1, (institution=="domesticr1"&role=="professor"&field=="math"))
vis_miss(restrictedData1)
```

One sees that 17.34% of the Math Sci Net citations data is missing. It appears there was some sort of systemtatic but unintentional error in the data collection from MathSciNet. We report 3 NaNs on A and B, 3 on A only, 0 on B and C, 50 on B Only, and 0 on C only. We manually checked the missing data and find that all but [Marta Civil](https://www.math.arizona.edu/~civil/) and [Jeffrey X Watt](https://science.iupui.edu/people/watt-jeffrey) (who are math educators) have Math Sci Net entries. The remaining omissions are fixed and we visually check for NaNs again.

```{r, warning=FALSE, echo = FALSE, include = FALSE}
table(restrictedData1$lettergroup[is.na(restrictedData1$amscit)])
```

```{r, echo = FALSE, include = FALSE}
df <- read_csv("./revised_data.csv")
df$age <- (2020- df$year)
df$citperyear <- df$citations/df$age
df$amscitperyear <- df$amscit/df$age
df$lettergroup <- as.factor(df$lettergroup)
df$gender <- as.factor(df$gender)
df$role <- as.factor(df$role)
df$highered <- as.factor(df$highered)
df$institution <- as.factor(df$institution)
df$research <- as.factor(df$research)
df$country <- as.factor(df$country)
df$security <- as.factor(df$security)
df$field <- as.factor(df$field)
df$simplefield <- as.factor(df$simplefield)
```

```{r, echo = FALSE}
restrictedData <- filter(df, (institution=="domesticr1"&role=="professor"&field=="math"))
vis_miss(restrictedData)
```

```{r, echo = FALSE, include = FALSE}
sum(is.na(restrictedData$amscit))
sum(is.na(restrictedData$amscitperyear))
nrow(restrictedData)
```
65/323 is empty for AMS citations per year. While visually the nans appear uniform, we will impose a stricter signifance level, say 2%, to assess the difference in AMS citations per year.

Now that we are comparing apples to apples, we reperform the main results.

## The Main Result of Paik-Rivin: R1 Math Professors Citations and Citations per Year

We will compare the mean number of citations and citations per year between signers of Letters A, B, and C. We will validate the significance of the difference between signers using a [permutation test](https://en.wikipedia.org/wiki/Resampling_(statistics)#Permutation_tests).

The following is copy and pasted from our paper (with edit in parentheses):

A permutation test is a nonparametric way of assessing the difference in mean between two populations. We are interested in whether an observed difference in mean is due to chance, and we can assess this in the following way.

0. Record the true difference in mean (or median) ($d\mu$).
1. $H_0: d\mu = 0, H_1: d\mu < 0$
2. Sample without replacement 1/2 of the combined data set (X) and what is left (Y)
3. Take the mean (or median) of X and Y and record the difference
4. Repeat 10,000 times and plot the histogram
5. Record the number of points (m) in the induced distribution that is more extreme than or equal to the observed $d\mu$. The probability m/10,000 is the probability that what was observed was due to chance.

```{r, echo = FALSE, include = FALSE}
# input data and the number of permutations
meanPermutation <- function(Data, n){
  output <- matrix(NA, ncol = 1, nrow = n)
  for(i in 1:n){
    #sample 1/2 of the data
    X_index <- sample(1:length(Data), floor(0.5 * length(Data)))
    Y_index <- setdiff(1:length(Data), X_index)
    X <- Data[X_index]
    Y <- Data[Y_index]
    #calculate the difference
    diff <- mean(X)-mean(Y)
    #store
    output[i, ] <- diff
  } 
  return(output)
}
```

```{r, echo = FALSE, include = FALSE}
# input data and the number of permutations
medianPermutation <- function(Data, n){
  output <- matrix(NA, ncol = 1, nrow = n)
  for(i in 1:n){
    #sample 1/2 of the data
    X_index <- sample(1:length(Data), floor(0.5 * length(Data)))
    Y_index <- setdiff(1:length(Data), X_index)
    X <- Data[X_index]
    Y <- Data[Y_index]
    #calculate the difference
    diff <- median(X)-median(Y)
    #store
    output[i, ] <- diff
  } 
  return(output)
}
```

### MathSciNet Citations for R1 Math Professors

```{r, warning=FALSE, echo = FALSE}
plot4 <- ggplot(df, aes(x=amscit))+
  geom_histogram(data=filter(df, ((lettergroup == "A Only"|lettergroup == "A and B")&(role=="professor"))),fill = "red", alpha = 0.2, binwidth = 1000) + 
  geom_histogram(data=filter(df, ((lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B")&(role=="professor"))),fill = "blue", alpha = 0.2, binwidth = 1000) +
  geom_histogram(data=filter(df,( (lettergroup == "C Only"|lettergroup == "B and C")&(role=="professor"))),fill = "green", alpha = 0.2, binwidth = 1000) +
  geom_vline(xintercept = mean(filter(df, ((lettergroup == "A Only"|lettergroup == "A and B")&(role=="professor")))$amscit, na.rm = TRUE), linetype="dotted", color = "red", size=1.5) +
  geom_vline(xintercept = mean(filter(df,( (lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B")&(role=="professor")))$amscit, na.rm = TRUE), linetype="dotted", color = "blue", size=1.5) +
  geom_vline(xintercept = mean(filter(df, ((lettergroup == "C Only"|lettergroup == "B and C")&(role=="professor")))$amscit, na.rm = TRUE), linetype="dotted", color = "green", size=1.5) +
  ggtitle("Math Sci Net Citations Comparison Professors (A=red), (B=blue), (C=green)")
plot4
```
```{r, echo = FALSE, include = FALSE}
mean(filter(df, ((lettergroup == "A Only"|lettergroup == "A and B")&(role=="professor")&(field=="math")))$amscit, na.rm = TRUE)
mean(filter(df, ((lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B")&(role=="professor")&(field=="math")))$amscit, na.rm = TRUE)
mean(filter(df, ((lettergroup == "C Only"|lettergroup == "B and C")&(role=="professor")&(field=="math")))$amscit, na.rm = TRUE)
median(filter(df, ((lettergroup == "A Only"|lettergroup == "A and B")&(role=="professor")&(field=="math")))$amscit, na.rm = TRUE)
median(filter(df, ((lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B")&(role=="professor")&(field=="math")))$amscit, na.rm = TRUE)
median(filter(df, ((lettergroup == "C Only"|lettergroup == "B and C")&(role=="professor")&(field=="math")))$amscit, na.rm = TRUE)
```

The mean number of citations for signers of letter A is 397 and the median is 261.
The mean number of citations for signers of letter B is 1435 and the median is 915.
The mean number of citations for signers of letter C is 2177 and the median is 1353.

The three hypotheses we would like to assess are:

1. $H_0: \mu(A) = \mu(B), H_1: \mu(A) < \mu(B)$

2. $H_0: \mu(B) = \mu(C), H_1: \mu(B) < \mu(C)$

3. $H_0: \mu(A) = \mu(C), H_1: \mu(A) < \mu(C)$

In the following figure, the vertical red bar is the observed difference for hypothesis 1 in the induced distribution, blue is the observed difference for hypothesis 2, and green is the observed difference for hypothesis 3.

```{r, echo=FALSE}
muA <- mean(filter(df, ((lettergroup == "A Only"|lettergroup == "A and B")&(role=="professor")))$amscit, na.rm = TRUE)
muB <- mean(filter(df, ((lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B")&(role=="professor")))$amscit, na.rm = TRUE)
muC <- mean(filter(df, ((lettergroup == "C Only"|lettergroup == "B and C")&(role=="professor")))$amscit, na.rm = TRUE)

val1 <- muA - muB
val2 <- muB - muC
val3 <- muA - muC

set.seed(0)
dist <- meanPermutation(na.omit(df$amscit),10000)

hist(dist,
     main = "Permutation Test on MSN cits only R1 Math Professors",
     xlab = "Differences in Mean")
abline(v=val1, col = "red")
abline(v=val2, col = "blue")
abline(v=val3, col = "green")

```
```{r, echo=FALSE, include = FALSE}
sum(dist < val1)/length(dist)
sum(dist < val2)/length(dist)
sum(dist < val3)/length(dist)
```

The induced p-value for hypothesis 1 is 0. The induced p-value for hypothesis 2 is 0.0002. The induced p-value for hypothesis 3 is 0. Hence we reject all three null hypotheses in favor of the alternative, and $\mu(A) < \mu(B) < \mu(C)$.

### MathSciNet Citations per Year for R1 Math Professors

Of course, we considered the fact that citations grow with age, so we calculated citations per year. There may be objections to this - one could hypothesize that citations per year grow with age - but we will soon thoroughly reject this claim.

```{r , warning=FALSE, echo = FALSE}
plot9 <- ggplot(df, aes(x=amscitperyear))+
  geom_histogram(data=filter(df, (lettergroup == "A Only"|lettergroup == "A and B")),fill = "red", alpha = 0.2, binwidth = 50) + 
  geom_histogram(data=filter(df, (lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B")),fill = "blue", alpha = 0.2, binwidth = 50) +
  geom_histogram(data=filter(df, (lettergroup == "C Only"|lettergroup == "B and C")),fill = "green", alpha = 0.2, binwidth = 50) +
  geom_vline(xintercept = mean(filter(df, (lettergroup == "A Only"|lettergroup == "A and B"))$amscitperyear, na.rm = TRUE), linetype="dotted", color = "red", size=1.5) +
  geom_vline(xintercept = mean(filter(df, (lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B"))$amscitperyear, na.rm = TRUE), linetype="dotted", color = "blue", size=1.5) +
  geom_vline(xintercept = mean(filter(df, (lettergroup == "C Only"|lettergroup == "B and C"))$amscitperyear, na.rm = TRUE), linetype="dotted", color = "green", size=1.5) +
  ggtitle("MathSciNet citperyear Comparison (A=red), (B=blue), (C=green)")
plot9
```

```{r, echo = FALSE, include= FALSE}
mean(filter(df, (lettergroup == "A Only"|lettergroup == "A and B"))$amscitperyear, na.rm = TRUE)
mean(filter(df, (lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B"))$amscitperyear, na.rm = TRUE)
mean(filter(df, (lettergroup == "C Only"|lettergroup == "B and C"))$amscitperyear, na.rm = TRUE)
median(filter(df, (lettergroup == "A Only"|lettergroup == "A and B"))$amscitperyear, na.rm = TRUE)
median(filter(df, (lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B"))$amscitperyear, na.rm = TRUE)
median(filter(df, (lettergroup == "C Only"|lettergroup == "B and C"))$amscitperyear, na.rm = TRUE)
```

The mean number of citations per year for signers of letter A is 16 and the median is 11.
The mean number of citations per year for signers of letter B is 41 and the median is 26.
The mean number of citations per year for signers of letter C is 56 and the median is 42.

The three hypotheses we would like to assess are:

1. $H_0: \mu(A_{citperyear}) = \mu(B_{citperyear}), H_1: \mu(A_{citperyear}) < \mu(B_{citperyear})$

2. $H_0: \mu(B_{citperyear}) = \mu(C_{citperyear}), H_1: \mu(B_{citperyear}) < \mu(C_{citperyear})$

3. $H_0: \mu(A_{citperyear}) = \mu(C_{citperyear}), H_1: \mu(A_{citperyear}) < \mu(C_{citperyear})$

As above, the vertical red bar is the observed difference for hypothesis 1 in the induced distribution, blue is the observed difference for hypothesis 2, and green is the observed difference for hypothesis 3.

```{r, echo = FALSE}
muA <- mean(filter(df, (lettergroup == "A Only"|lettergroup == "A and B"))$amscitperyear, na.rm = TRUE)
muB <- mean(filter(df, (lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B"))$amscitperyear, na.rm = TRUE)
muC <- mean(filter(df, (lettergroup == "C Only"|lettergroup == "B and C"))$amscitperyear, na.rm = TRUE)

val1 <- muA - muB
val2 <- muB - muC
val3 <- muA - muC

set.seed(0)
dist <- meanPermutation(na.omit(df$amscitperyear),10000)

hist(dist,
     main = "Permutation Test MSN citperyear R1 Math Professors",
     xlab = "Differences in Mean")
abline(v=val1, col = "red")
abline(v=val2, col = "blue")
abline(v=val3, col = "green")
```
```{r, echo=FALSE, include = FALSE}
sum(dist < val1)/length(dist)
sum(dist < val2)/length(dist)
sum(dist < val3)/length(dist)
```

The induced p-value for hypothesis 1 is 0. The induced p-value for hypothesis 2 is 0.0099. The induced p-value for hypothesis 3 is 0. Hence we reject all three null hypotheses, even assessing the induced p-value at a 2% significance level, in favor of the alternative, and we conclude that $\mu(A_{citperyear}) < \mu(B_{citperyear}) < \mu(C_{citperyear})$.

One may object of our usage of the mean here, opposed to the median. So we reperform the permutation test with the median. All three induced p-values are 0 so we can reject all three null hypotheses when comparing median citations per year.

```{r, echo=FALSE, include = FALSE}
muA <- median(filter(df, (lettergroup == "A Only"|lettergroup == "A and B"))$amscitperyear, na.rm = TRUE)
muB <- median(filter(df, (lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B"))$amscitperyear, na.rm = TRUE)
muC <- median(filter(df, (lettergroup == "C Only"|lettergroup == "B and C"))$amscitperyear, na.rm = TRUE)

val1 <- muA - muB
val2 <- muB - muC
val3 <- muA - muC

set.seed(0)
dist <- medianPermutation(na.omit(df$amscitperyear),10000)
```

```{r, echo=FALSE, include = FALSE}
sum(dist < val1)/length(dist)
sum(dist < val2)/length(dist)
sum(dist < val3)/length(dist)
```

## There is no evidence that Citations per Year grows with age

Let us check whether there is a relationship between age and citations per year in our limited dataset.

### 2020-01-22 index.Rmd
typo

---
title: "Response to Pachter’s Review"
author: "Joshua Paik and Igor Rivin"
date: "1/22/2020"
output:
  html_document:
    toc: yes
  pdf_document:
    keep_tex: yes
    toc: yes
---

@@

One sees that 17.34% of the Math Sci Net citations data is missing. It appears there was some sort of systemtatic but unintentional error in the data collection from MathSciNet. We report 3 NaNs on A and B, 3 on A only, 0 on B and C, 50 on B Only, and 0 on C only. We manually checked the missing data and find that all but [Marta Civil](https://www.math.arizona.edu/~civil/) and [Jeffrey X Watt](https://science.iupui.edu/people/watt-jeffrey) (who are math educators) have Math Sci Net entries. The remaining omissions are fixed and we visually check for NaNs again.

>>>>

---
title: "Response to Pachter’s Review"
author: "Joshua Paik and Igor Rivin"
date: "1/22/2020"
output:
  pdf_document:
    keep_tex: yes
    toc: yes
  html_document:
    toc: yes
---

@@

One sees that 17.34% of the Math Sci Net citations data is missing. It appears there was some sort of systematic but unintentional error in the data collection from MathSciNet. We report 3 NaNs on A and B, 3 on A only, 0 on B and C, 50 on B Only, and 0 on C only. We manually checked the missing data and find that all but [Marta Civil](https://www.math.arizona.edu/~civil/) and [Jeffrey X Watt](https://science.iupui.edu/people/watt-jeffrey) (who are math educators) have Math Sci Net entries. The remaining omissions are fixed and we visually check for NaNs again.
